The Anthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2023-04-14 paper.tex
rename ANNO-core/ANNO-domain in ANNOdc/ANNOds

\section{Ontological Architecture}\label{sec:architecture}
%Gesamtarchitektur, Zusammenhang zwischen GFO, ANNO-core, ANNO-domain, 3-Ontologien-Methode, alles mit Verweisen auf die nächsten Kapitel, vielleicht ein kleines Bild

@@

  \item ANNO-domain (\cref{sec:domain}) is a domain ontology for describing domain-specific entities to be used for annotating 
the parts of human skeleton (bones, teeth and landmarks, such as, mandible, left mandibular central secondary incisor tooth 
or mentale dexter) as well as to model their properties and relations (e.g., the distance between mentale dexter and mentale sinister) 
required to derive human phenotypes (e.g., sex or height). The ANNO-domain is embedded in ANNO-core 
(i.e., the ANNO-domain classes are subclasses of the ANNO-core classes).

  \item ANNO-core (\cref{sec:core}, \cref{fig:core}) is a core ontology to provide the core entities of a domain, such as, inter alia, 
the general anatomical categories (\enquote{Anatomical structure} and \enquote{Landmark}), categories for describing different characteristics of 
landmarks (\enquote{Landmark location}, \enquote{Landmark property} and \enquote{Landmark line region}) as well as 
the category \enquote{Phenotype} to model the rules for determining human phenotypes. The ANNO-core is founded by \emph{General Formal Ontology} (GFO)~\citep{gfo}.

@@

The \aw{} software (\cref{sec:aw}) is designed according to the three-ontology method~\citep{threeontologymethod}. 
This method is based on the interaction of a task ontology, a domain ontology and a top-level ontology. 
In our case, the ANNO-core plays the role of a task ontology and serves as the conceptual model of the software. 
One of the advantages of the three-ontology method is that the software only needs to implement the access to entities (classes, properties) 
of the task ontology (ANNO-core), whereas the entities of the corresponding domain ontology (ANNO-domain) are processed dynamically. 
The \aw{} uses the ANNO-core as an interface to access the ANNO-domain entities, e.g. by listing all subclasses of a specific ANNO-core class, such as \aurl{anno}{Bone}, as described in \cref{sec:aw}.

@@

\section{Description of ANNO-core}\label{sec:core}
% Definition/Beschreibung der Kern-Konzepte/Module und -Relationen mit Beispielen, GFO-Einbettung
\begin{figure}[h]
\includegraphics[width=\textwidth]{img/core.pdf}
\caption{ANNO-core ontology}\label{fig:core}
\end{figure}

@@

\todo{Umschreiben von Modulen auf ANNO-core}
\subsection{Functions}
Functions are categorized into sex determination and regress functions for body height estimation.
RDF is not optimized for mathematical formulas so we model those as literals.

@@

\section{Development of ANNO-domain}\label{sec:domain}
% Arbeit der Domänenexperten, Tabelle, OWL-Generator
\begin{figure}[h!t]
\includegraphics[width=\textwidth]{img/smog.png}
\caption{Excerpt of the spreadsheet-based input template used by the anthropologists.}\label{fig:smog}
\end{figure}

>>>>

\section{Ontological Architecture}\label{sec:architecture}
%Gesamtarchitektur, Zusammenhang zwischen GFO, ANNOdc, ANNOds, 3-Ontologien-Methode, alles mit Verweisen auf die nächsten Kapitel, vielleicht ein kleines Bild

@@

  \item ANNOds (\cref{sec:domain}) is a domain ontology for describing domain-specific entities to be used for annotating 
the parts of human skeleton (bones, teeth and landmarks, such as, mandible, left mandibular central secondary incisor tooth 
or mentale dexter) as well as to model their properties and relations (e.g., the distance between mentale dexter and mentale sinister) 
required to derive human phenotypes (e.g., sex or height). The ANNOds is embedded in ANNOdc 
(i.e., the ANNOds classes are subclasses of the ANNOdc classes).

  \item ANNOdc (\cref{sec:core}, \cref{fig:core}) is a core ontology to provide the core entities of a domain, such as, inter alia, 
the general anatomical categories (\enquote{Anatomical structure} and \enquote{Landmark}), categories for describing different characteristics of 
landmarks (\enquote{Landmark location}, \enquote{Landmark property} and \enquote{Landmark line region}) as well as 
the category \enquote{Phenotype} to model the rules for determining human phenotypes. The ANNOdc is founded by \emph{General Formal Ontology} (GFO)~\citep{gfo}.

@@

The \aw{} software (\cref{sec:aw}) is designed according to the three-ontology method~\citep{threeontologymethod}. 
This method is based on the interaction of a task ontology, a domain ontology and a top-level ontology. 
In our case, the ANNOdc plays the role of a task ontology and serves as the conceptual model of the software. 
One of the advantages of the three-ontology method is that the software only needs to implement the access to entities (classes, properties) 
of the task ontology (ANNOdc), whereas the entities of the corresponding domain ontology (ANNOds) are processed dynamically. 
The \aw{} uses the ANNOdc as an interface to access the ANNOds entities, e.g. by listing all subclasses of a specific ANNOdc class, such as \aurl{anno}{Bone}, as described in \cref{sec:aw}.

@@

\section{Description of ANNOdc}\label{sec:core}
% Definition/Beschreibung der Kern-Konzepte/Module und -Relationen mit Beispielen, GFO-Einbettung
\begin{figure}[h]
\includegraphics[width=\textwidth]{img/core.pdf}
\caption{ANNOdc ontology}\label{fig:core}
\end{figure}

@@

\todo{Umschreiben von Modulen auf ANNOdc}
\subsection{Functions}
Functions are categorized into sex determination and regress functions for body height estimation.
RDF is not optimized for mathematical formulas so we model those as literals.

@@

\section{Development of ANNOds}\label{sec:domain}
% Arbeit der Domänenexperten, Tabelle, OWL-Generator
\begin{figure}[h!t]
\includegraphics[width=\textwidth]{img/smog.png}
\caption{Excerpt of the spreadsheet-based input template used by the anthropologists.}\label{fig:smog}
\end{figure}
